NumPy---Array Computation for Python

### 2019-12-12 paper.tex
Minor edits

NumPy\cite{vanderwalt2011numpy} is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It underpins almost every library that does numerical
computation, including SciPy.  NumPy consists of the `ndarray`-object
along with utility functions to operate on such arrays.  Because of
its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy
array has become the {\it de facto} exchange format for array data in
Python.  That said, NumPy has now seen such widespread adoption that
not only the array object but also its {\it Application Programming
  Interface} (API) has become ubiquitous, and is imitated by, e.g.,
popular tensor computation libraries.

Over the past few years, with external financial support, NumPy has
seen renewed developer interest.  It is actively developed to meet the
needs of its community, including the addition of some fairly large
changes as outlined in Section XXX.

\section*{History}

NumPy was, at its inception, the unifier of the Python array
processing world.  Before NumPy, two packages—Numeric and
numarray—existed.  The first was developed by Paul DuBois and
colleagues at [the lab], and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed [by STSCI?]. In other words, these array
objects had different strengths and weaknesses, and therefore tended
to divide the already small user community.  In 200x, Travis Oliphant
wrote NumPy, a ``best of both worlds'' unification of the two
projects, based on Numeric.  The arrival of stable version 1.0 in 200x unified the scientific Python
community, and heralded the era of popular scientific computing in
Python that has led to formidable successes such as [cite some impacts here].

@@

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at X, mostly the package had rapid growth
and adoption without external investment.  In 2014, NumPy received its
first large grants from the Moore and Sloan foundations, totaling 1.3M
dollars.  This allowed for hiring several full time programmers and,
with the support of community members, mobilized a project that had
become stagnant for the several years prior.

@@

@@

\subsection*{Library organisation}
Broadly, the NumPy library consists of the following parts:
the NumPy array data structure; the so-called \emph{universal functions};
a set of library functions for manipulating arrays and doing scientific
computation; infrastructure libraries for unit tests and Python package
building; and the program \code{f2py} for wrapping Fortran code in Python.
The \code{ndarray} and the universal functions are generally considered
the core of the library.
In the following, we give a brief summary of these components of the
library.

@@

The \code{ndarray} is the data structure at the heart of NumPy.
The data structure stores regularly strided homogeneous data types
in a contiguous block memory, allowing for the efficient representation
of $n$-dimensional data. \fixme{[How much detail do we want here?]}
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

@@

@@

\section*{Recent technical improvements}

@@

\subsection*{Conclusion}

\subsubsection*{Impact Now}

NumPy, while having been wildly successful as a library and array
exchange project, has now also become a standard language for expressing array
and tensor computations.  For example, the NumPy API is implemented by
several popular tensor computation libraries including CuPy
(https://cupy.chainer.org/), Jax
(https://jax.readthedocs.io/en/latest/jax.numpy.html), PyTorch
(https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html),
and Apache MXNet
(https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html),
and to a lesser extent TensorFlow
(https://www.tensorflow.org/api\_docs/python/tf/Tensor).

@@

\subsubsection*{Future Development}

\section*{Data Availability}
% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.
All NumPy source code and most data generated for the current study are available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some supporting code and data have also been stored in other public repositories cited by this manuscript.

>>>>

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It underpins almost every library that does numerical
computation, including SciPy.  NumPy consists of the `ndarray`-object
along with utility functions to operate on such arrays.  Because of
its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy
array has become the {\it de facto} exchange format for array data in
Python.  That said, NumPy has now seen such widespread adoption that
not only the array object but also its {\it Application Programming
  Interface} (API) has become ubiquitous, and is imitated by, e.g.,
popular tensor computation libraries.

Over the past few years, with external financial support, NumPy has
seen renewed developer interest.  It is actively developed to meet the
needs of its community, including the addition of some fairly large
changes as outlined in Section~\ref{improvements}.

\subsection*{History}

NumPy was, at its inception, the unifier of the Python array
processing world.  Before NumPy, two packages—Numeric and
numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed at the Space Telescope Science Institute (STScI).
In other words, these array
objects had different strengths and weaknesses, and therefore tended
to divide the already small user community.  In 2005, Travis Oliphant
started NumPy, a ``best of both worlds'' unification of the two
projects, based on Numeric.  The arrival of stable version 1.0 in October
2006 unified the scientific Python
community, and heralded the era of popular scientific computing in
Python that has led to formidable successes such as [cite some impacts here].

@@

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
and adoption without external investment.  In 2014, NumPy received its
first large grants from the Moore and Sloan foundations, totaling 1.3M
dollars.  This allowed for hiring several full time programmers and,
with the support of community members, mobilized a project that had
become stagnant for the several years prior.

@@

\subsection*{Project goals / scope of library}

\subsection*{Current status (maturity, users)}

\subsection*{Enabled ecosystem }

% NEP 29 (which releases are supported in rolling window)

% NumPyDoc standard, recent improvements

@@

\subsection*{Library organization}
Broadly, the NumPy library consists of the following parts:
the NumPy array data structure; the so-called \emph{universal functions};
a set of library functions for manipulating arrays and doing scientific
computation; infrastructure libraries for unit tests and Python package
building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
The \code{ndarray} and the universal functions are generally considered
the core of the library.
In the following, we give a brief summary of these components of the
library.

@@

The \code{ndarray} is the data structure at the heart of NumPy.
The data structure stores regularly strided homogeneous data types
in a contiguous block memory, allowing for the efficient representation
of $n$-dimensional data.
More details about the data structure are given in ``The NumPy array:
a structure for efficient numerical computation.''\cite{vanderwalt2011numpy}.

@@

% Wheels build system

@@

\section*{Recent technical improvements}\label{improvements}

@@

% \subsubsection*{Impact Now}

NumPy, while having been wildly successful as a library and array
exchange project, has now also become a standard language for expressing array
and tensor computations.  For example, the NumPy API is implemented by
several popular tensor computation libraries including
CuPy\footnote{\url{https://cupy.chainer.org/}},
Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
and to a lesser extent
TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.

@@

% \subsubsection*{Future Development}

\section*{Data Availability}
% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.

All NumPy source code and most data generated for the current study are
available in the NumPy GitHub repository, \url{https://github.com/numpy}. Some
supporting code and data have also been stored in other public repositories
cited by this manuscript.
